---
title: "Bioc 2020 Tidytranscriptomics Solutions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solutions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Questions:  
1. What is the Fraction of Variance for PC1 and PC2? What do PC1 and PC2 represent?  
2. How many DE genes are there for treated vs untreated? What is the top DE gene by P value?  
3. What code can generate a heatmap of variable genes (starting from count_scaled)?  
4. What code can you use to visualise expression of the pasilla gene (gene id: FBgn0261552)  
5. What code can generate an interactive volcano plot that has gene symbols showing on hover?  
6. What code can generate a heatmap of the top 100 DE genes?

Suggested answers are below. You might have some different code e.g. to customise the volcano plot as you like. Feel free to comment on any of these solutions in the workshop website as described [here](https://github.com/stemangiola/bioc_2020_tidytranscriptomics/blob/master/CONTRIBUTING.md).

```{r out.width = "70%", message=FALSE, warning=FALSE}
# load libraries

# tidyverse core packages
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)

# tidyverse-friendly packages
library(tidyHeatmap)
library(tidybulk)
library(ggrepel)
library(plotly)

# load data
data("pasilla", package = "bioc2020tidytranscriptomics")

# create tidybulk tibble
counts_tt <- pasilla %>% 
  tidybulk()

# scale counts
counts_scaled <- counts_tt %>% scale_abundance(factor_of_interest = condition)

# create density plots
counts_scaled %>%
  filter(!lowly_abundant) %>%
	pivot_longer(cols = c("counts", "counts_scaled"), names_to = "source", values_to = "abundance") %>%
  ggplot(aes(x=abundance + 1, group=sample, color=condition)) +
	geom_density() +
	facet_wrap(~source) +
	scale_x_log10() +
	theme_bw()
```

1. What is the Fraction of Variance for PC1 and PC2? 

```{r}
counts_scal_PCA <-
  counts_scaled %>%
  reduce_dimensions(method="PCA")
```

Answer: PC1: 47%, PC2: 25%

What do PC1 and PC2 represent?

```{r out.width = "70%"}
counts_scal_PCA %>%
	pivot_sample() %>%
	ggplot(aes(x=PC1, y=PC2, colour=condition, shape=type)) + 
  geom_point() +
  geom_text_repel(aes(label=sample), show.legend = FALSE) +
	theme_bw()
```

Answer: PC1 represents variance due to treatment effect(treated vs untreated). PC2 represents variance due to sequencing type single vs paired.


```{r}
counts_de <-
  counts_tt %>%
  test_differential_abundance(.formula = ~ 0 + condition + type, 
                              .contrasts = c("conditiontreated - conditionuntreated"), 
                              omit_contrast_in_colnames = TRUE)
```

2. How many DE genes are there for treated vs untreated (FDR < 0.05)?

```{r}
counts_de %>% 
  filter(significant == TRUE) %>% 
  summarise(num_de = n_distinct(feature))
```

Answer: 1128

What is the top DE gene by P value? 

```{r}
topgenes <- counts_de %>%
	pivot_transcript() %>%
  arrange(PValue) %>%
  head(6)

topgenes
```	

Answer: FBgn0025111


3. What code can generate a heatmap of variable genes (starting from count_scaled)?  

```{r out.width = "70%"}
counts_scaled %>% 
  
  # filter lowly abundant
  filter(!lowly_abundant) %>%
	
	# extract 500 most variable genes
	keep_variable( .abundance = counts_scaled, top = 500) %>%
	
	# create heatmap
	heatmap(
	      .column = sample,
	      .row = feature,
	      .value = counts_scaled,
	      annotation = c(condition, type),
	      transform = log1p 
	  )
```

4. What code can you use to visualise expression of the pasilla gene (gene id: FBgn0261552) 

```{r out.width = "70%"}
counts_scaled %>%
	
	# extract counts for pasilla gene
	filter(feature == "FBgn0261552") %>%
	
	# make stripchart
	ggplot(aes(x = condition, y = counts_scaled + 1, fill =condition, label = sample)) +
	geom_boxplot() +
	geom_jitter() +
	scale_y_log10()+
	theme_bw()
```

5. What code can generate an interactive volcano plot that has gene ids showing on hover?  

```{r out.width = "70%"}
p <- counts_de %>%
	pivot_transcript() %>%

  # Subset data
	filter(!lowly_abundant) %>%
	mutate(significant = FDR<0.05 & abs(logFC) >=2) %>%

  # Plot
	ggplot(aes(x = logFC, y = PValue, label=feature)) +
	geom_point(aes(color = significant, size = significant, alpha=significant)) +
	geom_text_repel() +
	
	# Custom scales
	scale_y_continuous(trans = "log10_reverse") +
	scale_color_manual(values=c("black", "#e11f28")) +
	scale_size_discrete(range = c(0, 2)) +
	theme_bw()

ggplotly(p, tooltip = c("text"))
```
Tip: You can use "text" instead of "label" if you don't want the column name to show up in the hover e.g. above will give "FBgn0261552" rather than "feature:FBgn0261552".



6. What code can generate a heatmap of the top 100 DE genes?

```{r out.width = "70%"}
top100 <- 
	counts_de %>%
	pivot_transcript() %>%
	arrange(PValue) %>%
	head(100)

counts_scaled %>% 
  filter(feature %in% top100$feature) %>%
	heatmap(
	      .column = sample,
	      .row = feature,
	      .value = counts_scaled,
	      annotation = c(condition, type),
	      transform = log1p 
	  )
```
